The \emph{GlobalMinimum} (MINIMUM) library of Jonas Mockus (1989)
implements some of the earliest practical Bayesian global-optimization
methods---one-step Bayesian search over a simple surrogate, Wiener-process
one-dimensional models, low-discrepancy designs, and clustering
multistart---that anticipate the acquisition-function machinery of modern
Bayesian optimization. We modernize this Fortran~66/77 codebase along the
two standard routes: \emph{wrapping} the original compiled routines behind
idiomatic R and Python packages, and \emph{translating} the core methods to
Rust. The exercise yields three contributions. First, we release polished R
and Python packages in which every optimizer runs the genuine 1989
algorithms (with callback objectives, deterministic seeding, full
evaluation traces, and input validation that retrofits memory- and
I/O-safety onto COMMON-block Fortran), plus a Rust crate exposing both the
translation and the wrapped original. Second, we dissect a
translation-fidelity incident: the Rust port of the one-step Bayesian
planner was up to $12\times$ slower than the wrapped Fortran---an artifact
initially misattributed to foreign-function overhead. Controlled
decomposition shows call overhead is microseconds and near-identical across
backends; the real cause was two dropped pruning rules (a partial-distance
early exit and a second-minimum bound) whose restoration recovers bitwise
trajectory agreement \emph{and} speed parity. The same audit surfaced
transposed coefficient matrices in re-implemented Shekel and Hartmann test
functions---quiet correctness bugs of the rewrite route. Third, we
benchmark the 1989 methods against modern global optimizers in both
ecosystems (SciPy, NLopt, CMA-ES, Gaussian-process Bayesian
optimization, GenSA, DEoptim, GA) on a COCO-style
shifted-instance suite of 16~test families at dimensions 2--10 under
matched evaluation budgets, using performance and data profiles, with
external validation on the official BBOB/COCO suite; the wrapped
methods replicate bitwise across the two ecosystems. On the classical
test set the coordinate Wiener-model search EXKOR is competitive with
the best modern methods (solving 56\% of instances to gap $10^{-2}$,
statistically tied with SHGO, dual annealing, and GenSA); on BBOB,
whose functions are predominantly non-separable and ill-conditioned,
it drops to mid-field (15\% versus 36\% for dual annealing), cleanly
quantifying how much of that strength is separability. At tight
budgets modern annealing and Gaussian-process expected improvement
dominate the 1989 acquisition machinery decisively---the latter at
three orders of magnitude more solver-internal time per run. Finally,
we ask whether the historic components compose: EXKOR/A2, which routes
EXKOR through the eigenframe estimated by the library's own screening
routine, repairs four of EXKOR's five zero-solve families and improves
its mean rank in both ecosystems, at the cost of a statistical tie in
aggregate solve rate---composition buys robustness, not throughput.
All code, data, and scripts are available for replication.
