---
title: "globalopt: the GlobalMinimum library in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{globalopt: the GlobalMinimum library in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The `globalopt` package compiles the original *GlobalMinimum* Fortran
library of Bayesian and other global optimization methods (Jonas Mockus,
1989) directly into R.  The methods predate — and in several cases
introduced — the ideas behind modern Bayesian optimization: surrogate
models, expected-improvement-style acquisition, and low-discrepancy
initial designs.

```{r setup}
library(globalopt)
```

## Quick start

Minimize the built-in `furasn` objective (a Rastrigin-like separable
function with global minimum $-2$ at the origin) over a box:

```{r}
a <- c(-1, -1)
b <- c(1, 1)
r <- bayes1(a, b, evaluations = 200, initial_points = 20, objective = "furasn")
r
```

The result contains the full evaluation trace:

```{r}
head(r$points)
plot(r$values, type = "l", xlab = "evaluation", ylab = "objective")
```

## R objectives and compiled objectives

The `objective` argument accepts either an R function or the *name* of a
compiled built-in test objective.  With a built-in name the optimization
loop never re-enters R, which makes runs of cheap objectives many times
faster — useful for benchmarking the methods themselves:

```{r}
f <- function(x) sum((x - 0.3)^2)
bayes1(a, b, 100, 20, f)$best_f

if (requireNamespace("microbenchmark", quietly = TRUE)) {
  microbenchmark::microbenchmark(
    r_callback = bayes1(a, b, 200, 20, furasn),
    compiled   = bayes1(a, b, 200, 20, "furasn"),
    times = 5
  )
}
```

## Methods

| Function | Origin (1989 Fortran) | Type |
| --- | --- | --- |
| `bayes1()` | BAYES1 | One-step Bayesian global method |
| `mig1()`, `mig2()` | MIG1, MIG2 | Monte-Carlo baselines |
| `lbayes()` | LBAYES | Bayesian stochastic local search |
| `unt()` | UNT | Multi-start with Wiener-model uncertainty |
| `glopt()` | GLOPT | Clustering global method |
| `extr()`, `exkor()` | EXTR, EXKOR | 1-D Wiener-model search and its coordinate version |
| `lpmin()` | LPMIN | LP-tau (Sobol-type) deterministic search |
| `mivar4()` | MIVAR4 | Variable-metric local method |
| `flexi()` | FLEXI | Constrained flexible simplex |
| `reqp()` | REQP | Recursive quadratic programming |
| `anal1()`, `anal2()` | ANAL1, ANAL2 | Variable screening |

Every function runs the original compiled Fortran by default
(`backend = "fortran"`).  For the core methods a pure-R reference
implementation is available via `backend = "reference"` — useful for
understanding the algorithms and for validating the bindings:

```{r}
bayes1(a, b, 20, 20, "furasn")$values[1:5]
bayes1(a, b, 20, 20, furasn, backend = "reference")$values[1:5]
```

## Determinism and seeding

The library draws random points from a 15-word lagged-Fibonacci generator
(`ATS`) whose state the package resets before every run, so results are
deterministic by default and reproduce a fresh-process run of the 1989
code.  Pass an integer `seed` to run perturbed replications:

```{r}
mig2(a, b, 100, "furasn")$best_f          # canonical run
mig2(a, b, 100, "furasn", seed = 1)$best_f
mig2(a, b, 100, "furasn", seed = 2)$best_f
```

## Constrained optimization

`flexi()` and `reqp()` support equality and inequality constraints via a
constraint function returning equalities first (feasible at 0), then
inequalities (feasible when non-negative):

```{r}
# minimize furasn subject to x1 >= 0.05
con <- function(x) x[1] - 0.05
reqp(c(0.2, 0.2), imax = 50, constraints = con, n_ineq = 1,
     objective = "furasn")$best_x
```

## Variable screening

The analysis routines estimate which variables (or pairs) drive the
objective, from any design of evaluated points:

```{r}
r <- mig2(c(-1, -1, -1), c(1, 1, 1), 100,
          function(x) x[1]^2 + 0.1 * x[2]^2 + 0 * x[3])
anal2(c(-1, -1, -1), c(1, 1, 1), r$points, r$values)$influence
```

## Limits inherited from the 1989 code

The original working arrays impose limits that the package validates
before calling Fortran: dimension at most 20 for most methods (100 for
`mig1`, `lbayes`, `mivar4`, `reqp`), at most 1000 evaluations for
`mig2`/`bayes1` and 500 for `unt`/`extr`/`exkor`.  The code keeps state in
COMMON blocks, so calls are not thread-safe; do not call these functions
concurrently within one process.

## References

Mockus, J. (1989). *Bayesian Approach to Global Optimization: Theory and
Applications*. Kluwer Academic Publishers.

Mockus, J., Tiesis, V., Zilinskas, A. (1978). The application of Bayesian
methods for seeking the extremum. *Towards Global Optimization*, 2, 117-129.
